---
title: "Review"
author: "Alan T. Arnholt"
date: "3/3/2017"
output: 
  bookdown::html_document2:
    highlight: textmate
    theme: yeti
---


```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(comment = NA, fig.show = 'as.is', fig.align = 'center', fig.height = 7, fig.width = 7, prompt = TRUE, highlight = TRUE, tidy = FALSE, warning = FALSE, message = FALSE, tidy.opts=list(blank = TRUE, width.cutoff= 75, cache = TRUE))
```

# Body Temperatures

```{r}
BT <- read.csv("http://www.lock5stat.com/datasets1e/BodyTemp50.csv")
library(dplyr)
BT %>%
  summarize(mean(BodyTemp), sd(BodyTemp), n())
library(ggplot2)
ggplot(data = BT, mapping = aes(x = BodyTemp)) + 
  geom_dotplot(fill = "lightblue") + 
  theme_bw()
```

## Null and Alternative Hypothesis

\begin{equation}
H_O: \mu = 98.6 \\
H_A: \mu \neq 98.6
(\#eq:nullalt)
\end{equation}

**Note:** In order to construct a **sampling distribution** to assess the evidence in `BT`, one must generate new samples consistent with the null hypothesis $(\mu = 98.6)$.  One way to generate samples that are consistent with the null hypothesis is to add 0.34 to all of the values in `BodyTemp`.

```{r}
BT <- BT %>%
  mutate(NBT = BodyTemp + 0.34) 
head(BT)
summarize(BT, mean(NBT))
```

### Generate bootstrap distribution

```{r}
B <- 10000 - 1
xbarstarA <- numeric(B)
xbarstar <- numeric(B)
for(i in 1:B){
  BSA <- sample(BT$NBT, size = 50, replace = TRUE)
  BS <- sample(BT$BodyTemp, size = 50, replace = TRUE)
  xbarstarA[i] <- mean(BSA)
  xbarstar[i] <- mean(BS)
}
mean(xbarstarA)
mean(xbarstar)
sd(xbarstarA)
sd(xbarstar)
ggplot(data = data.frame(x = xbarstarA), aes(x = x)) + 
  geom_density(fill = "pink") + 
  theme_bw() + 
  geom_vline(xintercept = 98.26)
p_value <- (sum(xbarstarA <= 98.26) + 1)/(B + 1)
p_value
ggplot(data = data.frame(x = xbarstar), aes(x = x)) + 
  geom_density(fill = "pink") + 
  theme_bw()
```

### Two bootstrap CIs

#### Percentile CI

```{r}
PCI <- quantile(xbarstar, probs = c(0.025, 0.975))
PCI
```

#### Bootstrap SE

```{r}
CIB <- mean(BT$BodyTemp) + c(-1, 1)*qnorm(.975)*sd(xbarstar)
CIB
```

## Classical Sampling Distribution

**Assume** $X \sim\mathcal{N}(98.6, 0.77) \rightarrow \bar{X}\sim\mathcal{N}(98.6, 0.77/\sqrt{50} = `r 0.77/sqrt(50)`)$

### Simulation approach

```{r}
SIMS <- 10000 - 1
xbarC <- numeric(SIMS)
xbarA <- numeric(SIMS)
for(i in 1:SIMS){
  xbarC[i] <- mean(rnorm(50, 98.26, 0.77))
  xbarA[i] <- mean(rnorm(50, 98.26 + 0.34, 0.77))
}
xbarS <- mean(xbarC)
SES <- sd(xbarC)
c(xbarS, SES)
# Simulation
ggplot(data = data.frame(x = xbarA), aes(x = x)) + 
  geom_density(fill = "lightblue") + 
  theme_bw()
p_value <- (sum(xbarA <= 98.26) + 1)*2/(SIMS + 1)
p_value
CIS <- quantile(xbarC, probs = c(0.025, 0.975))
CIS
# Theoretical
p_value <- pnorm(98.26, 98.6, 0.77/sqrt(50))*2
p_value
CIT <- mean(BT$BodyTemp) + c(-1, 1)*qnorm(0.975)*0.77/sqrt(50)
CIT
## Or
PASWR2::z.test(x = BT$BodyTemp, sigma.x = 0.77, mu = 98.6)
```

## Data Tweaking

```{r}
BT <- BT %>%
  mutate(Gender = factor(Gender, levels = c(0, 1), labels = c( "Male", "Female")))
DT::datatable(BT)
BT %>%
  group_by(Gender) %>%
  summarize(mean(BodyTemp), sd(BodyTemp), n())
```

\begin{equation}
H_O: \mu_{\text{Female}} - \mu_{\text{Male}} = 0 \\
H_O: \mu_{\text{Female}} - \mu_{\text{Male}} \neq 0 
(\#eq:nullalt2)
\end{equation}

### Classical Approach

```{r}
t.test(BodyTemp ~ Gender, data = BT)
STUFF <- BT %>%
  group_by(Gender) %>%
  summarize(MA = mean(BodyTemp))
obsdiff <- diff(tapply(BT$BodyTemp, BT$Gender, mean))
obsdiff
```

### Randomization Approach

```{r}
SIMS <- 10^4 - 1
diffmeans <- numeric(SIMS)
for(i in 1:SIMS){
  diffmeans[i] <- diff(tapply(BT$BodyTemp, sample(BT$Gender), mean))
}
mean(diffmeans)
sd(diffmeans)
pvalue <- ((sum(diffmeans >= obsdiff) + 1)/(SIMS + 1))*2
pvalue
```

### Bootstrap CI

```{r}
B <- 10^4 - 1
bsd <- numeric(B)
for(i in 1:B){
  bsm <- sample(BT$BodyTemp[BT$Gender == "Male"], size = 25, replace = TRUE)
  bsf <- sample(BT$BodyTemp[BT$Gender == "Female"], size = 25, replace = TRUE)
  bsd[i] <- mean(bsf) - mean(bsm)
}
CI <- quantile(bsd, probs = c(0.025, 0.975))
CI
```